---
title: "Zac's Gamma-Poisson Analysis"
format: html
editor: visual
---

## How many vehicles will be stolen in Chicago tomorrow?

In this section we will break down our statistical modelling for determining the number of stolen vehicles in Chicago for a new day.

```{r}
library(tidyverse)
library(bayesrules)
library(lubridate)
library(ggplot2)
```

First we read the data set containing all currently reported vehicle thefts since the start of the 2026 year.

```{r}
df <- read.csv("../data/chicago_vehicle_theft_crimes_2026.csv")
# mdy_hms handles the string formatting "mm/dd/yyyy hh:mm:ss AM" automatically
df$Date <- mdy_hms(df$Date)
# Convert the date to a numeric date since the start of the dataset
df$day_numeric <- as.numeric(difftime(df$Date, min(df$Date), units = "days"))
# Convert the numeric dates to discrete numbers 
# Remove the incomplete current day
df <- df %>% mutate(day_n = floor(day_numeric)) %>% filter(day_n != 52)
```

Then grouping the reports by day and counting them, allows us to create a visualization to judge for any noticeable trends to model.

```{r}
# Aggregate the counts of reported theft cases
daily_counts <- df %>% group_by(day_n) %>% summarize(theft_count = n())
# Plot the counts over days to looks for correlation
ggplot(daily_counts, aes(x = day_n, y = theft_count)) +
  geom_line(color = "steelblue") +  # Line shows the trend
  geom_point(color = "darkblue") +  # Points show individual days
  labs(title = "Daily Frequency of Vehicle Thefts",
       x = "Day Number",
       y = "Count of Thefts") +
  theme_minimal()
```

Generally from the plot there is no specific trend from day-to-day which makes using an independence assumption more plausible.

Plotting the histograms of the reported numbers will show if mostly the numbers are centered around a similar value.

```{r}
# Treating all days as independent events we can construct a density plot
ggplot(daily_counts, aes(x = theft_count)) +
  geom_histogram(fill = "steelblue", color = "black", binwidth = 3) +
  labs(title = "Distribution of Daily Theft Counts",
       x = "Number of Thefts per Day",
       y = "Counts") +
  theme_minimal()
```

Looking at the plot there is relatively high variance in the daily reported number of stolen vehicles in Chicago, however the number is generally centered around the 40-50 range, indicating that on average days have similar numbers of cars stolen.

To model these counts we propose discrete random variables that range from 0 to infinity, which are independent and have the same mean. One type of random variable that can model these conditions is the Poisson random variable

$ \text{Let } Y_i\text{ represent the number of car thefts in on a day }i:$

$$Y_i|\lambda \sim Poisson(\lambda)$$

Furthermore the rate of the Poisson random variable can be modeled as a Gamma random variable.

$ \text{Let }\lambda \text{ represent the daily car theft rate}: $

$$\lambda \sim Gamma(\alpha, \beta)$$

From research into priors the annual car theft rate in Chicago is 8.25 for every 1000 people. Given that Chicago has a population of around 2.72 million, that leads to 22440 total stolen cars per year in Chicago!

Since we are using the independent and identically distributed assumption for the car theft rate in Chicago, we expect that the expected rate should be close to one 365th of the annual rate, which leads to a rate of 61.44.

For the variance of the rate, the monthly 5 year counts range from 630 to 3186 which translates into an average daily rate range of 21 to 106 and since each month in a 5 year range accounts for one 60th of the data, I suggest that choosing a distribution that has tail probabilities around 0.0166, for either or possibly both tails. However a Gamma distribution is not flexible enough to model all three constraints simultaneously so I decided to only use the left tail probability assumption. In other words the 0.0166th quantile of the prior Gamma distribution should be 21. $\text{From our assumption that the average rate should be 61.44: }$ $$\text{E}(\lambda) = \frac{\alpha}{\beta}= 61.44 $$

$$\beta= \frac{\alpha}{61.44}$$

$
\text{From the assumption about the 0.0166th quantile being 21:}
$

$$
\text{P}(\lambda \le 21) = 0.0166
$$

$
\text{Substituting the CDF of a Gamma distribution:}
$

$$
\text{P}(\lambda \le x) = F(x; \alpha, \beta) =\int_{0}^{x} \frac{\beta^\alpha}{\Gamma(\alpha)} t^{\alpha-1} e^{-\beta t} \, dt =   \frac{\gamma(\alpha, \beta x)}{\Gamma(\alpha)}
$$

$ \text{Substituting the contraint from the expectation of a Gamma distribution:} $

$$ \text{P}(\lambda \le 21) =\frac{\gamma(\alpha, \frac{21}{61.44}\alpha)}{\Gamma(\alpha)} = 0.0166$$

$ \text{Analytically solving this is difficult so we computationally solve the unique solution for }\alpha :$

$$
\text{P}(\lambda \le 21) = F(21,\alpha,\frac{\alpha}{61.44}) =  \text{pgamma}(21, \alpha,\frac{\alpha}{61.44}) = 0.0166
$$

$
\text{Equivalently solving the single root of the funtion } g(\alpha):
$

$$  g(\alpha) = \text{pgamma}(21, \alpha,\frac{\alpha}{61.44}) - 0.0166= 0 $$

In R we can use the uniroot function to computationally solve the root for us:

```{r}
# Targets for gamma
target_mean <- 61.44
target_tail_v <- 21
target_tail_p <- 1/60

# Defining g(alpha)
g <- function(alpha) {
  beta <- alpha / target_mean
  tail_p <- pgamma(target_tail_v, shape = alpha, rate = beta)
  return(tail_p - target_tail_p)
}

# Solving the root of g
solution <- uniroot(g, lower = 0.01, upper = 100)

# Substituting in the values for alpha and beta
prior_alpha <- solution$root
prior_beta <- prior_alpha / target_mean

print(paste("Alpha:", round(prior_alpha, 4)))
print(paste("Beta:", round(prior_beta, 4)))
```

Visualizing the constructed prior:

```{r}
plot_gamma(prior_alpha, prior_beta)
```

Checking our initial constraint of a mean rate of 61.44

```{r}
# Get the mean of the distribution
summarize_gamma(prior_alpha,prior_beta)$mean
```

and a high enough variance to make rates with smaller than the minimum average rates to not account for more than 0.0166 probability.

```{r}
# Get the quantiles of the minimum and maximum average rates in the past 5 years
pgamma(c(21,106),prior_alpha,prior_beta)
```

Notice that this prior has constrained lower rates more than higher rates, but could have been reversed if there is more belief that rates that are higher than the maximum rate in the past 5 years should be less likely.

Using a gamma-poisson model allows us to make use of a conjugate prior:

```{r}
total = sum(daily_counts$theft_count)
days = length(daily_counts$day_n)
plot_gamma_poisson(prior_alpha,prior_beta, total, days)
```

resulting in the following posterior:

```{r}
alpha <- prior_alpha + total
beta <- prior_beta + days
plot_gamma(alpha, beta)
```

```{r}
summarize_gamma_poisson(prior_alpha, prior_beta, total, days)
```

Using this we can construct a 95% credible interval for the mean rate for tomorrows car theft rate

```{r}
qgamma(c(0.025,0.975), alpha, beta)
```

We can also simulate the actual number stolen tomorrow and construct the predictive interval

```{r}
set.seed(84735)
# Simulate 500,000 potential average theft rates
n_sims <- 500000
lambda_sims <- rgamma(n_sims, shape = alpha, rate = beta)

# For each lambda, simulate a number of thefts
predictive_thefts <- rpois(n_sims, lambda = lambda_sims)
quantile(predictive_thefts, c(0.025, 0.975))

```

With our assumptions, there is a 95% probability that the number of cars stolen tomorrow will be between 33 and 59 inclusive.

```{r}
predictive_df <- data.frame(thefts = predictive_thefts)

ggplot(predictive_df, aes(x = thefts)) +
  geom_bar(aes(y = after_stat(count / sum(count))), 
           fill = "forestgreen", 
           color = "black") +
  geom_vline(xintercept = mean(predictive_df$thefts), 
             color = "red", 
             linetype = "dashed") +
  labs(title = "Posterior Predictive Distribution of Daily Thefts",
       subtitle = "Red dashed line represents the mean predicted daily count",
       x = "Predicted Number of Thefts",
       y = "Probability") +
  theme_minimal()
```

The constructed posterior predictive distribution seems to be centered at a mode and have slight right skew seen by how the mean is to the right of the mode.

If we wanted to predict the single best number for the number of cars stolen tomorrow the mode of the distribution would be the best estimate.

```{r}
counts <- count(predictive_df, thefts) %>% mutate(p = n/n_sims)
counts[which.max(counts$n),]
```

The most likely value is 45 with a probability of 0.058756.

```{r}
summary(predictive_df)
```

On average we should see the rate of 45.57 cars stolen per night and a 50% chance of having less than or equal to 45 cars stolen.
