---
title: "Chicago Motor Vehicle Theft Analysis"
author: "Thuy, Trian, Puru, Zac"
format: revealjs
---

## Are vehicles in Chicago susceptible targets of crime?

-   Nationally 1 million vehicles are stolen annually.
-   On average 3 out of 1000 people will have their car stolen this year
-   People are skeptical about leaving things in the car, but what about the car itself?
-   Chicago being one of the most crime ridden cities in the US makes a good worst-case benchmark for the nation

## Bullets

When you click the **Render** button a document will be generated that includes:

-   Content authored with markdown
-   Output from executable code

## Code

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

## 1. Exploring Times of Vehicle Theft

What is the probability that a vehicle theft in Chicago occurs at night? (night = between 6pm and 6am)

::: {.fragment}
```{r}
library(readr)
library(bayesrules)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)

df <- read.csv("../data/chicago_vehicle_theft_crimes_2026.csv")

# add a column converting Date to DateTime
df <- df %>%
  mutate(DateTime = strptime(Date, format = "%m/%d/%Y %I:%M:%S %p"))

df %>%
  mutate(hour = hour(DateTime),
         period = ifelse(hour >= 18 | hour < 6, "Night", "Day")) %>%
  count(hour, period) %>%
  ggplot(aes(x = factor(hour), y = n, fill = period)) +
  geom_col()+
  labs(
    title = "2026 Vehicle Thefts by Hour of Day",
    x = "Hour of Day",
    y = "Count"
  )
```
:::

## Prior

- Source: FBI National Incident-Based Reporting System
- 52.4% out of 750,000 vehicle thefts in 2024 occurred between 6pm and 6am
$$\pi \sim Beta(104, 96)$$

## Likelihood

```{r echo=TRUE}
nrow(df)
```

```{r echo=TRUE}
count_night <- df %>%
  filter(hour(DateTime) >= 18 | hour(DateTime) < 6) %>%
  nrow()

count_night
```
- Likelihood: y ~ Binomial(n, π)
- n = 2369, y = 1371

## Posterior Approximation of π
```{r}
grid_data <- data.frame(pi_grid = seq(from = 0.50, to = 0.61, length = 101))
grid_data <- grid_data %>% 
  mutate(prior = dbeta(pi_grid, 104, 96),
         likelihood = dbinom(1371,2369, pi_grid))

grid_data <- grid_data %>% 
  mutate(unnormalized = likelihood * prior,
         posterior = unnormalized / sum(unnormalized))

ggplot(grid_data, aes(x = pi_grid, y = posterior)) + 
  geom_point() + 
  geom_segment(aes(x = pi_grid, xend = pi_grid, y = 0, yend = posterior))
```

## Posterior Simulation of π

```{r}
set.seed(84735)

post_sample <- sample_n(grid_data, size = 10000, 
                        weight = posterior, replace = TRUE)
set.seed(84735)
ggplot(post_sample, aes(x = pi_grid)) + 
  geom_histogram(aes(y = after_stat(density)), color = "white", bins=34) + 
  stat_function(fun = dbeta, args = list(1475, 1094)) + 
  lims(x = c(0.5, 0.61))
```
```{r echo=TRUE}
mean(post_sample$pi_grid)
quantile(post_sample$pi_grid, c(0.025, 0.975))
```

## Conclusion

$\pi | y = 0.574$

The probability that a vehicle theft occurs at night is approximately 0.574

## Fighting Crime with Bayesian Analysis

(delete this later) Who will these results help in the future and how? 

1. Help people protect their personal vehicles by being aware of dangerous times to leave vehicles unlocked or out. Help law enforcement enforce regulations to limit vehicle theft, such as increasing street lighting.

## Limitations + Next Steps

(delete this later) If you had all the resources in the world and unlimited time what would you do on this topic? What are the limitations of the analysis you have done that you could correct in the future with unlimited resources?

1. No detailed insight outside of probability that a vehicle will be stolen at night. Want an hour-to-hour analysis.